The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts

### 2020-05-13 write-up/covid-writeup.tex
update covid paper

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{setspace}
\DeclareGraphicsExtensions{.eps, .ps}
\usepackage{soul,color,hyperref}
\usepackage{amsmath, amsthm, amsfonts}
% \usepackage[margin=1.5in]{geometry}

@@

This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the long-forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.  First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.  This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.  We show that estimates of the effective reproduction number may be similarly impacted.  Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.  Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.  We then end the paper with a discussion of the benefits of randomized testing, use of auxiliary information is most relevant, and the distinction between data quality and decision making in the current pandemic.

@@

Here we highlight the core insights from~\cite{Meng2018} that are relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
$$
\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 \right].
$$
This points to a troubling and paradoxical situation: the error relative to SRS increases as a function of population size.  \cite{Meng2018} termed this the ``Law of Large Populations'' (LLP).  This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $E_{\I} [\rho_{I,Y} ]$ equal) can yield wildly different estimates due to population size.  Large countries like the US may have similar true prevalence rates as smaller countries like the UK.  Even under similar testing strategies, the mean-squared error in the prevalence rate for the US will have substantially more variation in comparison to SRS.  Comparing the US to the UK, for example, the MSE will increase by a factor of almost five.  Thus conclusions drawn from observed case-count records may be not just wrong, but very wrong.

@@

Tests are imperfect.  COVID-19 testing is no exception. Here we investigate the interplay between imperfect testing and selection bias.  In discussions of inaccurate testing, the standard assumption is measurement-error leads to parameter attenuation.  When paired with selection bias, however, the two sources of error become entangled, and resulting biases can become magnified, muted, or even switch signs.

First we require some additional notation.  Let $P_j$ be an indicator of measurement error, equal to $1$ when we mismeasure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.  Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.  We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
$$
\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
$$
where $Z = 1-2Y$. The first term represents the perfect testing regime, the second term represents the interaction between imperfect testing and selection bias, while the third time represents the bias due to imperfect testing.  From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement-error.  In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.

We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement-error and selection bias.  To do this, we define the sampling rates differential.  Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 1)$ be the sampling rates.  Then $\Delta = f_1 - f_0$ is the sampling rate differential.  Using these terms, we can re-express the first two terms as
$$
\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
$$
The final term in brackets is the \emph{measurement-error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement-error interaction with prevalence and sampling rates interaction with prevalence. Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the appendix.  This implies the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.

@@

Consider again the current COVID-19 pandemic. We continue to assume the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.  In Section~\ref{section:est_dq}, we discuss recent research suggesting a false negative rate around 17.2\% and a false positive rate around 0.05\%. Under these rates and the current US prevalence rates, the ratio of the MSE to MSE under no measurement error is 0.436; if we switch false negative rates down to 5\% and increase the false positive rate to 5\% then the relative MSE is 2.89.  This is merely to demonstrate that in some cases we see a huge increase in MSE and in other settings we have a huge decrease in MSE.  What drives this is the false negative rate interaction with prevalence and sampling rates.  Therefore, whether we are better or worse off with respect to the MSE is a very difficult question to answer.

The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling. We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms. In the appendix, we show that the error now can be expressed as
\begin{equation}
\label{eq:error}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
\end{equation}
The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement-error.  The second term is the interaction between selection bias and measurement error.  Figure~\ref{fig:heatmap} presents the measurement-error data quality adjustment as a function of the relative frequency and odds ratio.  We see that the adjustment can be both positive and negative as well as greater than or less than zero.

@@

\begin{wrapfigure}{r}{0.5\textwidth}
\centering
\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.15$ and $FN=0.10$. Color scaled so blue = $-6$, white = 0, and red = $6$.}
\label{fig:heatmap}
\vspace{-0.3cm}
\end{wrapfigure}

@@

Unfortunately, ratio estimators do not cancel errors as now both the numerator and the denominator are uncertain.  This means selection bias and measurement error can have paradoxical effects.  Here we consider the ratio estimators for the relative change in the prevalence rate.  We let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).  Then the ratio estimator is given by $r = \bar y_t / \bar y_{t-1}$.  We assume both numerator and denominator are the prevalence estimates adjusted for measurement-error. If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_t/\bar y_{t-1} = y_t/y_{t-1}$).  We wish to express the error $\frac{\bar y_t}{\bar y_{t-1}} - \frac{\bar Y_{t}}{\bar Y_{t-1}}$ in statistical terms.  Using a Taylor series approximation, we show in the appendix that the error can be expressed approximately as
$$
\begin{aligned}
\frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
&\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
\end{aligned}
$$
where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.  The magnitude of the error depends on the population-level ratio $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate. The second term represents the ``cancellation'' that the naive analyst is hoping will occur.  The analyst's intuition only holds when the difference is equal to zero.  This occurs at times when data quality, sampling fraction, measurement-error, and prevalence are all constant across time.

@@

Figure \ref{fig:ratio-bias} shows a trajectory of the true ratio and the potential biased estimators under an SIR model for the epidemic dynamics, with state evolution given by the following ODE
$$
\begin{aligned}
\frac{\partial S_t}{\partial t} &= - \beta S_t I_t N^{-1}; \quad
\frac{\partial I_t}{\partial t} = \beta S_t I_t N^{-1} - \gamma I_t; \quad
\frac{\partial R_t}{\partial t} = - \gamma I_t \\
\end{aligned}
$$
where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total poulation size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.   Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$.  In each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.  From a decision making perspective, such biases have potential impact.  First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.  Of course, the analysis supports the argument that the estimates based on available data are overly pessimistic. What both sides miss is that the direction of bias is non-constant over time.  After we hit the peak, estimates will likely under-estimate prevalence.  Such bias puts pressure on lawmakers and governors to prematurely relax social distancing measures.  Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''  It appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.  Note, however, that the standard errors on such estimates will be incredibly wide due to the low effective sample size as discussed in Section~\ref{section:casecount}.

\subsection{Estimation of effective reproduction number}
\label{section:r0-estimation}
Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.  We show here how these estimates are also impacted by selection bias and measurement error.  For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates. \cite{Bettencourt2008} show that under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived.  In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19~\citep{Sanche2020}.  Using this, a simple estimator of the effective reproduction number at time $t$ can be given by
$$
R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
$$
Of course, we do not observe $K_t$ and $K_{t-1}$ but noisy proxies $y_t$ and $y_{t-1}$, i.e., the observed case-count on day $t$ and $t-1$ respectively. The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$. However, if we think about estimating number $K_t$ under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \bar y_t$.  Unfortunately the number of individuals susceptible on day $t$ is unknown.  Here, we assume that $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \bar y_t / \bar y_{t-1} \right)$ be the estimated effective reproduction number on day $t$.
This estimator adjusts for sample size variation across days but not the varying population size. The estimator is close in spirit to the estimator derived under assuming the process is fully observed.  Similar to Section~\ref{section:rates}, the proxy has error in both numerator and denominator that may cause issues.  We again wish to express the statistical error of $\hat R_t - R_t$ in useful terms.  We can re-arrange the error decomposition from Section~\ref{section:rates} to show that this error is given by
$$
\begin{aligned}
\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \bigg] \\
&\times \left[ 1 - \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t) \right] \bigg) - \frac{1}{\gamma} \log \left( \frac{S_t}{S_{t-1}} \right).
\end{aligned}
$$
So here we see the exact same trade-off as with the rates but on the logarithmic scale.  For small values of $x$, we have $\log(1+x) = x$ and so the biases are similar.  However, the error is no longer scaled by the rate of change in prevalence $\bar Y_2/\bar Y_1$. Instead the error is scaled by the serial interval $\gamma^{-1} = 1/7$.  Moreover, the error depends not on aggregate prevalence $\bar Y_t$ but fraction of new cases $\bar K_t$. The final term is the error due to varying population sizes.  Since $S_t \leq S_{t-1}$, the final term $- \log (S_t/S_{t-1}) \geq 0$. Figure~\ref{fig:r0-bias} displays the bias for the effective reproductive number and how it is impacted by the relative sampling fraction.  Here for simplicity, we assume the fraction of the population that is susceptible remains large and near the total population size.  Under these assumption, the potential bias is equivalent to that for the ratio estimator adjusted to be on the logarithm scale.  We note that changes to this assumption can drastically impact the potential bias.

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../methods/figs/sir_ratio.png}
  \caption{Ratio estimator}
  \label{fig:ratio-bias}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt.png}
  \caption{Effective reproductive rate estimator}
  \label{fig:r0-bias}
\end{subfigure}
\caption{Potential bias due to self-selection and measurement-error from a deterministic SIR model with $\beta = 1.4$ and $\gamma = 0.2$.  Here we assume $f = 0.02$, $FP = 0.01$, $FN = 0.15$, and use a range of relative sampling fraction $M = f_1/f_0$.}
\label{fig:rates}
\end{figure}

@@

In such settings, simple random sampling is doomed to fail; however, there are potential solutions.  In particular, one can stratify the population into risk categories (based on population density, occupation, age, and other important factors) and perform Neyman allocation~\citep{Cochran77}.  That is, sample individuals at random in strata $h$ with size proportional to $P_h \sqrt{\bar Y_h \times (1-\bar Y_h)}$ where $P_h$ is the strata frequency in the population and $\bar Y_h$ the prevalence in the strata. Of course, $P_h$ and $\bar Y_h$ are not known a priori; however, if we can get decent estimates, we can see huge benefits in settings where the prevalence varies widely by strata. One could go attempt to go further and incorporate contact network information.  However, such data is difficult to obtain in practice.  Indeed, geographic location, occupation, age and basic demographic information are much easier to obtain and are a good proxy for the latent network connectivity.

>>>>

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{setspace}
\DeclareGraphicsExtensions{.eps, .ps}
\usepackage{hyperref}
\usepackage{amsmath, amsthm, amsfonts}
% \usepackage[margin=1.5in]{geometry}

@@

This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the too often forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.  First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.  This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.  We show that estimates of the effective reproduction number may be similarly impacted.  Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.  Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.  We then end the paper with a discussion of the benefits of randomized testing, use of auxiliary information to improve over simple random sampling, and the distinction between data quality and decision making in the current pandemic.

@@

Here we highlight two key insights from~\cite{Meng2018} that are relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
$$
\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 \right].
$$
This points to a troubling and paradoxical situation: the error relative to SRS increases as a function of population size.  \cite{Meng2018} termed this the ``Law of Large Populations'' (LLP).  This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $E_{\I} [\rho_{I,Y} ]$ equal) can yield wildly different estimates due to population size.  Large countries like the US may have similar true prevalence rates as smaller countries like the UK.  Even under similar testing strategies, the mean-squared error in the prevalence rate for the US will have substantially more variation in comparison to SRS.  Comparing the US to the UK, for example, the MSE will increase by a factor of almost five.  Thus conclusions drawn from observed case-count records may be not just wrong, but very wrong.

@@

Tests are imperfect.  COVID-19 testing is no exception. Here we investigate the interplay between imperfect testing and selection bias.  In discussions of inaccurate testing, the standard assumption is measurement-error leads to parameter attenuation.  When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.

First we require some additional notation.  Let $P_j$ be an indicator of measurement error, equal to $1$ when we mismeasure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.  We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
$$
\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
$$
where $Z = 1-2Y$; see Appendix~\ref{app:imperfect} for the derivation. The first term represents the perfect testing regime, the second term represents the interaction between imperfect testing and selection bias, while the third time represents the bias due to imperfect testing.  From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement-error.  In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.

We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement-error and selection bias.  To do this, we define the sampling rates differential.  Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 1)$ be the sampling rates.  Then $\Delta = f_1 - f_0$ is the sampling rate differential.  Using these terms, we can re-express the first two terms as
$$
\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
$$
The final term in brackets is the \emph{measurement-error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement-error interaction with prevalence and sampling rates interaction with prevalence. Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the appendix, which implies the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.

@@

Consider again the current COVID-19 pandemic. For now, we continue to assume the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.  In Section~\ref{section:est_dq}, we discuss recent research suggesting a false negative rate around 17.2\% and a false positive rate around 0.05\%. Under these rates and the current US prevalence rates, the ratio of the MSE to MSE under no measurement error is 0.436; if we switch false negative rates down to 5\% and increase the false positive rate to 5\% then the relative MSE is 2.89.  This is merely to demonstrate that in some cases we see a huge increase in MSE and in other settings we have a huge decrease in MSE.  What drives this is the false positive and negative rate interaction with prevalence and sampling rates.  Therefore, whether we are better or worse off with respect to the MSE is a very difficult question to answer.

The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling. We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms. In the appendix, we show that the error now can be expressed as
\begin{equation}
\label{eq:error}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
\end{equation}
The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement-error.  The second term is the interaction between selection bias and measurement error.  Figure~\ref{fig:heatmap} presents the measurement-error data quality adjustment as a function of the relative frequency and odds ratio.  We see that the adjustment can be both positive and negative as well as a range of magnitudes.

@@

\begin{wrapfigure}{r}{0.5\textwidth}
\centering
\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
\label{fig:heatmap}
\vspace{-0.3cm}
\end{wrapfigure}

@@

Unfortunately, ratio estimators do not cancel errors as now both the numerator and the denominator are uncertain.  This means selection bias and measurement error can have paradoxical effects.  Here we consider the ratio estimators for the relative change in the prevalence rate.  We let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence (across the entire population) at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).  Then the ratio estimator is given by $r = \bar y_t / \bar y_{t-1}$.  We assume both numerator and denominator are the prevalence estimates adjusted for measurement-error. If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_t/\bar y_{t-1} = y_t/y_{t-1}$).  We wish to express the error $\frac{\bar y_t}{\bar y_{t-1}} - \frac{\bar Y_{t}}{\bar Y_{t-1}}$ in statistical terms.  Using a Taylor series approximation, we show in the appendix that the error can be expressed approximately as
$$
\begin{aligned}
\frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
&\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
\end{aligned}
$$
where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.  The magnitude of the error depends on the population-level ratio $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate. The second term represents the ``cancellation'' that the naive analyst is hoping will occur.  The analyst's intuition only holds when the difference is equal to zero.  This occurs at times when data quality, sampling fraction, measurement-error, and prevalence are all constant across time.

@@

Figure \ref{fig:ratio-bias} shows a trajectory of the true ratio and the potential biased estimators under an SIR model for the epidemic dynamics, with state evolution given by the following ODE
$$
\begin{aligned}
\frac{\partial S_t}{\partial t} &= - \beta S_t I_t N^{-1}; \quad
\frac{\partial I_t}{\partial t} = \beta S_t I_t N^{-1} - \gamma I_t; \quad
\frac{\partial R_t}{\partial t} = - \gamma I_t \\
\end{aligned}
$$
where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total poulation size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.   Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$ (black line).  In each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.  From a decision making perspective, such biases have potential impact.  First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.  Of course, the analysis supports the argument that the estimates based on available data are overly pessimistic. What both sides miss is that the direction of bias is non-constant over time.  After we hit the peak, estimates will likely under-estimate prevalence.  Such bias puts pressure on lawmakers and governors to prematurely relax social distancing measures.  Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''  It appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.  Note, however, that the standard errors on such estimates will be incredibly wide due to the low effective sample size as discussed in Section~\ref{section:casecount}.
\label{section:r0-estimation}
Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.  We show here how these estimates are also impacted by selection bias and measurement error.  For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates. \cite{Bettencourt2008} show that under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived.  In particular, under an SIR model, the number of case counts on day $t$ is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval which is approximately $7$ days for COVID-19~\citep{Sanche2020}.  Using this, a simple estimator of the effective reproduction number at time $t$ can be given by
$$
R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
$$
Of course, we do not observe $K_t$ and $K_{t-1}$ but noisy proxies $y_t$ and $y_{t-1}$, i.e., the observed case-count on day $t$ and $t-1$ respectively. The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$ where here $y_t$ isc conceptualized as the number of new cases on day $t$. However, if we think about estimating number $K_t$ under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \bar y_t$.  Unfortunately the number of individuals susceptible on day $t$ is unknown.  Here, we assume that $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \bar y_t / \bar y_{t-1} \right)$ is the estimated effective reproduction number on day $t$.
This estimator adjusts for sample size variation across days but not the varying population size. The estimator is close in spirit to the estimator derived under assuming the process is fully observed.  Similar to Section~\ref{section:rates}, the proxy has error in both numerator and denominator that may cause issues.  We again wish to express the statistical error of $\hat R_t - R_t$ in useful terms.  We can re-arrange the error decomposition from Section~\ref{section:rates} to show that this error is given by
$$
\begin{aligned}
\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
&\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \frac{1}{\gamma} \log \left( \frac{S_t}{S_{t-1}} \right).
\end{aligned}
$$
So here we see the exact same trade-off as with the rates but on the logarithmic scale.  For small values of $x$, we have $\log(1+x) = x$ and so the biases are similar.  However, the error is no longer scaled by the rate of change in prevalence $\bar Y_2/\bar Y_1$. Instead the error is scaled by the serial interval $\gamma^{-1} = 1/7$.  Moreover, the error depends not on aggregate prevalence $\bar Y_t$ but fraction of new cases $\bar K_t$. The final term is the error due to varying population sizes.  Since $S_t \leq S_{t-1}$, the final term $- \log (S_t/S_{t-1}) \geq 0$. Figure~\ref{fig:r0-bias} displays the bias for the effective reproductive number and how it is impacted by the relative sampling fraction.  Here for simplicity, we assume the fraction of the population that is susceptible remains large and near the total population size.  Under these assumption, the potential bias is similar to that for the ratio estimator adjusted to be on the logarithm scale.  Changes to this assumption can impact the potential bias.  The key difference between Figure~\ref{fig:ratio-bias} and Figure~\ref{fig:r0-bias} is that prevalence depends on the fraction of the population infected at time $t$ while $\hat R_t$ depends on the fraction of new cases in the population. This leads to differences in when the bias is most pronounced.

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../methods/figs/sir_ratio.png}
  \caption{Ratio estimator}
  \label{fig:ratio-bias}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../methods/figs/sir_rt.png}
  \caption{Effective reproductive rate estimator}
  \label{fig:r0-bias}
\end{subfigure}
\caption{Potential bias due to self-selection and measurement-error from a deterministic SIR model with $\beta = 1.4$ (black) and $\gamma = 0.2$ (red).  Here we assume $f = 0.02$, $FP = 0.01$, $FN = 0.15$, and use a range of relative sampling fraction $M = f_1/f_0$.}
\label{fig:rates}
\end{figure}

@@

In such settings, simple random sampling is doomed to fail; however, there are potential solutions.  In particular, one can stratify the population into risk categories (based on population density, occupation, age, and other important factors) and perform Neyman allocation~\citep{Cochran77}.  That is, sample individuals at random in strata $h$ with size proportional to $P_h \sqrt{\bar Y_h \times (1-\bar Y_h)}$ where $P_h$ is the strata frequency in the population and $\bar Y_h$ the prevalence in the strata. Of course, $P_h$ and $\bar Y_h$ are not known a priori; however, if decent estimates can be obtained, we can see huge benefits in settings where the prevalence varies widely by strata. One could go attempt to go further and incorporate contact network information.  However, such data is difficult to obtain in practice.  Indeed, geographic location, occupation, age and basic demographic information are much easier to obtain and are a good proxy for the latent network connectivity.
